Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-05-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean up repo

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. TODO: add stratification. \\
\hline
\end{tabular}
\end{table}

@@

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset"). And add overview of possible predictor matrix values.]

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

**TODO: add paragraph about novice and more experienced readers.**

@@

$$
\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
$$

```{r include=FALSE}
# data
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
  mutate(school = SCHOOL,
         popular = POPULAR,
         sex = SEX,
         texp = TEXP,
         .keep = "none")
popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
```

@@

```{r}
popmis <- popmis[, c("school", "popular", "sex", "texp")] 
```
Plot the missing data pattern:

@@

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.

 <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. -->

Plot the correlations in the incomplete data:

@@

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.

@@

```{r pop-hist}
ggmice(popmis, aes(texp)) +
  geom_histogram(fill = "white") +
  facet_grid(. ~ is.na(popular), labeller = label_both)
```

@@

### Complete case analysis (not recommended)

Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.

```{r pop-cca, echo=FALSE}
est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)

```

@@

The first imputation model that we'll use is likely to be invalid. We do *not* use the cluster identifier `school` as imputation model predictor. With this model, we ignore the multilevel structure of the data, despite the high ICC. This assumes exchangeability between units. We include it purely to illustrate the effects of ignoring the clustering in our imputation effort. We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).

@@

TODO: remove the broom.mixed output, use mitml only

@@

TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.

TODO: Add that the more the random effects are of interest, the more you need ml models.

@@

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) # TODO: add random effects ct?
```

>>>>

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. TODO: add stratification. \\
\hline
\end{tabular}
\end{table}

@@

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset"). Add that the more the random effects are of interest, the more you need ml models. And add overview of possible predictor matrix values.]

@@

**TODO: add where novice and more experienced readers should start reading.** For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

@@

<!-- $$ -->
<!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->
<!-- $$ -->
```{r}
mod <- popular ~ 1 + sex + texp + sex:texp + (1 + sex | school)
```

```{r include=FALSE}
# data
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
  mutate(school = SCHOOL,
         popular = POPULAR,
         sex = SEX,
         texp = TEXP,
         .keep = "none")

est_true <- lme4::lmer(mod, popcomp) %>% broom.mixed::tidy()

popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
```

@@

```{r}
popmis <- popmis[, c("school", "popular", "sex", "texp")] 
```

Plot the missing data pattern:

@@

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}.

 <!-- The ICC in the incomplete data is `round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.-->

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the observed values of `popular` are related to observed values of other variables. Plot the pair-wise complete correlations in the incomplete data:

@@

This shows us that both `sex` and `texp` may be useful imputation model predictors. Moreover, the missingness in `popular` may depend on the observed values of other variables. We'll highlight one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.

@@

```{r pop-hist}
ggmice(popmis, aes(texp)) +
  geom_histogram(fill = "white") +
  facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
```

@@

<!-- ### Complete case analysis (not recommended) -->

<!-- Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.  -->

<!-- ```{r pop-cca, echo=FALSE} -->
<!-- est_cca <- lme4::lmer(mod, popmis) %>% broom.mixed::tidy() -->
<!-- # as.matrix(est_cca$estimate) - as.matrix(est_true) -->
<!-- est_cca$estimate - est_true$estimate -->
<!-- ``` -->

@@

The first imputation model that we'll use is likely to be invalid. We do *not* use the cluster identifier `school` as imputation model predictor. With this model, we ignore the multilevel structure of the data, despite the high ICC. This assumes exchangeability between units. We include it purely to illustrate the effects of ignoring the clustering in our imputation effort. 
<!-- We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).  -->

@@

**TODO: remove the broom.mixed output, use mitml only**

@@

**TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.**

@@

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
```

### 2022-05-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Clean up repo

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset''). And add overview of
possible predictor matrix values.{]}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.

\textbf{TODO: add paragraph about novice and more experienced readers.}

@@

\[
\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
\]

@@

The missingness is univariate and sporadic, which is illustrated in the
missing data pattern in Figure \ref{fig:pop_pat}. This tells us that the
multilevel structure of the data should probably be taken into account.
If we don't, we'll may end up with incorrect imputations, biasing the
effect of the clusters towards zero.

Plot the correlations in the incomplete data:

@@

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight one other variable
to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{texp})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{texp}
separately for the pupils with known popularity and missing popularity.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> ggmice(popmis, aes(texp)) +
+   geom_histogram(fill = "white") +
+   facet_grid(. ~ is.na(popular), labeller = label_both)
\end{CodeInput}

@@

\hypertarget{complete-case-analysis-not-recommended}{%
\subsubsection{Complete case analysis (not
recommended)}\label{complete-case-analysis-not-recommended}}

Complete case analysis ignores the observations with missingness
altogether, which lowers statistical power and may even introduce bias
in MCAR situations.

@@

The first imputation model that we'll use is likely to be invalid. We do
\emph{not} use the cluster identifier \texttt{school} as imputation
model predictor. With this model, we ignore the multilevel structure of
the data, despite the high ICC. This assumes exchangeability between
units. We include it purely to illustrate the effects of ignoring the
clustering in our imputation effort. We'll use the default imputation
methods in \texttt{mice()} (predictive mean matching to impute the
continuous variables and logistic regression to impute binary
variables).

@@

TODO: remove the broom.mixed output, use mitml only

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     3.354     0.245    13.704     6.831     0.000     3.260     0.813 
sex             1.247     0.332     3.759     6.154     0.009     4.160     0.849 
texp            0.110     0.017     6.320     6.084     0.001     4.286     0.852 
sex:texp       -0.027     0.022    -1.225     5.914     0.267     4.631     0.862

                            Estimate 
Intercept~~Intercept|school    0.177 
sex~~sex|school                0.232 
Intercept~~sex|school         -0.043 
Residual~~Residual             0.670 
ICC|school                     0.209

@@

TODO: Add that it doesn't work with syst missing (only sporadically).
There's some pro's and con's. May not differ much if the number of
clusters is low.

TODO: Add that the more the random effects are of interest, the more you
need ml models.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) # TODO: add random effects ct?
\end{CodeInput}
\end{CodeChunk}

>>>>

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset''). Add that the more the
random effects are of interest, the more you need ml models. And add
overview of possible predictor matrix values.{]}

@@

\textbf{TODO: add where novice and more experienced readers should start
reading.} For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod <- popular ~ 1 + sex + texp + sex:texp + (1 + sex | school)
\end{CodeInput}
\end{CodeChunk}

@@

The missingness is univariate and sporadic, which is illustrated in the
missing data pattern in Figure \ref{fig:pop_pat}.

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the observed values of
\texttt{popular} are related to observed values of other variables. Plot
the pair-wise complete correlations in the incomplete data:

@@

This shows us that both \texttt{sex} and \texttt{texp} may be useful
imputation model predictors. Moreover, the missingness in
\texttt{popular} may depend on the observed values of other variables.
We'll highlight one other variable to illustrate, but ideally one would
inspect all relations. The questions we'll ask are: `Does the missing
data of pupil popularity (\texttt{popular}) depend on observed teacher
popularity (\texttt{texp})?'. This can be evaluated statistically, but
visual inspection usually suffices. We'll make a histogram of
\texttt{texp} separately for the pupils with known popularity and
missing popularity.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> ggmice(popmis, aes(texp)) +
+   geom_histogram(fill = "white") +
+   facet_grid(. ~ is.na(popular), scales = "free", labeller = label_both)
\end{CodeInput}

@@

@@

The first imputation model that we'll use is likely to be invalid. We do
\emph{not} use the cluster identifier \texttt{school} as imputation
model predictor. With this model, we ignore the multilevel structure of
the data, despite the high ICC. This assumes exchangeability between
units. We include it purely to illustrate the effects of ignoring the
clustering in our imputation effort.

@@

\textbf{TODO: remove the broom.mixed output, use mitml only}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     3.489     0.179    19.542     9.227     0.000     1.927     0.714 
sex             1.138     0.197     5.776    11.588     0.000     1.424     0.644 
texp            0.100     0.014     7.339     6.914     0.000     3.177     0.809 
sex:texp       -0.019     0.015    -1.217     7.708     0.260     2.576     0.773

                            Estimate 
Intercept~~Intercept|school    0.123 
sex~~sex|school                0.143 
Intercept~~sex|school          0.014 
Residual~~Residual             0.656 
ICC|school                     0.158

@@

\textbf{TODO: Add that it doesn't work with syst missing (only
sporadically). There's some pro's and con's. May not differ much if the
number of clusters is low.}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
\end{CodeInput}
\end{CodeChunk}
